An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-09-20 nips_2016.tex
Updated the paper draft with around a third of my handwritten comments. Still a lot to go.

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}

@@

\author{
  John Canny \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{canny@berkeley.edu}
  \And
  Haoyu Chen \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{haoyuchen@berkeley.edu}
  \And
  Daniel Seita \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{seita@berkeley.edu}
  \And
  Xinlei Pan \\
  Department of Bioengineering \\
  University of California, Berkeley \\
  \texttt{xinleipan@berkeley.edu}
  \And 
  Biye Jiang \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{bjiang@berkeley.edu}
 }

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to perform
Metropolis-Hastings tests on the full data. Previous work on reducing
the cost of Metropolis-Hastings tests yield variable data consumed per
sample, with only constant factor reductions vs. using the full
dataset for each sample.  Here we present a method that can be tuned
to provide arbitrarily small batch sizes, by adjusting either proposal
step size or temperature. Our approach uses the natural noise present
in minibatch likelihood esimates to furnish the randomness in a
Metropolis-Hastings test. Our test uses the noise-tolerant Barker
acceptance test with a novel additive correction variable.  The
resulting test can be combined with minibatch proposals to yield
updates with the same complexity as a simple SGD update. In this paper
we derive the test, analyze its performance, discuss its
implementation, and present several experiments.
\end{abstract}

@@

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. We are interested primarily
in large dataset applications, where the goal is to sample a posterior
distribution $p(\theta \mid x_1, \ldots, x_N)$ of parameter $\theta$,
and where the number of data instances $N$ is large.  The
Metroplis-Hastings method (M-H) generates sample candidates from a
proposal distribution $q$ which is in general different from the
target distribution $p$, and decides whether to accept or reject them
based on an acceptance test. The acceptance test is usually a
Metropolis test~\cite{Metropolis1953,hastings70}. Conventional
Metropolis-Hastings requires all $N$ data instances to generate one
posterior sample.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and have high sample efficiency. They also
map very well onto hardware such as GPUs. In contrast, M-H requires
calculations over the full dataset to produce a new sample.  Recent
results from~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets
(minibatches) of the full dataset. The tests depend on minibatch
statistics, and on the value of an additional random variable $u$. The
amount of data consumed for each test varies significantly from one
minibatch to the next, and depends on the current sample, the proposed
sample, and on the random variable $u$. By contrast,
\cite{conf/uai/MaclaurinA14} performs exact tests but requires a lower
bound on parameter distribution across its domain. The amount of data
reduction depends on the accuracy of this bound, and such bounds are
only available for relatively simple distributions.

Here we derive a new test which incorporates the variability in
minibatch statistics as {\em a natural part of the test}. Because of
this, the amount of data required for each test is fixed with high
probability and in expected value. We use a Barker test function
\cite{Barker65} rather than a Metropolis test, which makes the test
naturally error tolerant. The idea of using a noise-tolerant test
using Barker's test function were suggested but not explored
empirically in \cite{Bardenet15} section 6.3. But the asymptotic test
statistic CDF and the Barker function are different, which leads to
fixed errors for the approach in \cite{Bardenet15}. Here we show that
the difference between the distributions can be corrected with an
additive random variable. This leads to a test which is fast, and
whose error can be made arbitrarily small.

@@

\item Increase the minibatch size. log acceptance variance scales as $1/k$ vs the
  minibatch size $k$. Increased minibatch size also reduces the error rate for
  the test.

@@

Its worth discussing at this point what are the goals (typically) of
M-H sampling on very large datasets.  By the Bernstein-von Mises
theorem, the posterior distribution of the parameter $\theta$ for a
Bayesian inference task is asymptotically normal, and has variance
that scales inversely with the number of data samples $N$. This mode
is extremely sharp for large datasets, which may contain millions or
billions of samples. Simply sampling from this distribution is one
application, but an efficient proposal distribution
\cite{OptimalScaling01} has similar variance to the target
distribution and will diffuse to it extremely slowly from an
initialization value which is (likely to be) many standard deviations
away. If there are any other strong modes, it is very likely for the
sampler to find one of them and become trapped in it when run at the
normal distribution temperature (T=1). A common solution is to anneal
the sampler, running first at high temperature (scaling log
likelihoods by 1/T) which flattens the likelihood landscape.  This is
turn reduces the variance of the log acceptance probability and allows
our acceptance test to be applied.

A second question concerns step size. Once we have fixed temperature,
our variance constraint implies that we have to trade-off proposal
step size $s$ and batch size $b$ ($b \propto p^2$). i.e. we can make
many small steps, or one large step, with a given batch of data. One
of the primary drivers of this work is our belief in the value of
small steps. For applications to neural networks or other models where
the posterior is multimodel, posterior inference is arguably a search
process. Covering the search space densely with small steps is much
more valuable than few sparse steps toward the nearest optimum. In
this mode, Metropolis-Hastings can be used in similar fashion to
Stochastic Gradient Descent. The goal in SGD is to make gradual
progress to a posterior mode with each step, taking small steps so
that the cumulative displacement has progressively lower variance. A
substantial part of the computational work of MCMC on massive datasets
will be similarly in reaching a stationary distribution, which really
means finding a deep posterior mode. Taking noisy small steps will
nevertheless make steady progress to a posterior mode since their bias
is in that direction. We demonstrate this behavior with our experiments.

@@

\begin{itemize}
\item We develop a new, more efficient (in samples per test) minibatch acceptance test with quantifiable error bounds. The test
  uses a novel additive correction variable to implement a Barker test based on minibatch mean and variance. 
\item We analyze the test for accuracy and speed.
\item We compared performance of our new test and prior approaches on several datasets. We demonstrate the test is several orders
  of magnitude more efficient than prior work measured as data consumed per test, and that it does not suffer
  from long-tailed batch sizes (up to the dataset size). 
\end{itemize}

@@

In the Metropolis-Hastings method~\cite{gilks1996markov,brooks2011handbook}, a difficult-to-compute
probability distribution $p(\theta)$ is sampled using a Markov chain $\theta_1,\ldots,\theta_n$. The 
sample $\theta_{t+1}$ at time $t+1$ is generated using a candidate $\theta'$ from a (simpler) proposal
distribution $q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test is usually
a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
\alpha(\theta_t,\theta') = \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)} \wedge 1
\end{equation}
where $a \wedge b$ denotes $\min(a,b)$.  With probability $\alpha(\theta_t,\theta')$, we accept $\theta'$
and set $\theta_{t+1} = \theta'$, otherwise set
$\theta_{t+1}=\theta_t$.  The test is often implemented with an
auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison $u < \alpha(\theta_t,\theta')$. $\mathcal{U}(a,b)$ denotes the uniform distribution on the interval $[a,b]$. 
For notational simplicity, from now on we will drop the subscript $t$ for the current sample $\theta_t$ and
denote is simply as $\theta$.

The acceptance test guarantees detailed balance, which means
\begin{equation}\label{detailed_balance1}
  p(\theta)p(\theta'\mid\theta) = p(\theta')p(\theta \mid\theta')
\end{equation}
where $p(\theta'\mid\theta)$ is the probability of a transition from state $\theta$ to $\theta'$. Here
$p(\theta'\mid\theta) = q(\theta'\mid\theta)\alpha(\theta,\theta')$ so the detailed balance equation becomes
\begin{equation}\label{detailed_balance2}
  p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
\end{equation}
The detailed balance condition, together with ergodicity, guarantee that the
Markov chain has a unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution for $\theta$ based on some
observed data $x_1,\ldots,x_N$. i.e. we want to sample from 
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}
\alpha(\theta,\theta') = 
 \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
\theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)} \wedge 1
\end{equation}
where $p_0(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the
probabilities of the observations. Computing samples this way requires
the use of all $N$ training data points, but this is very expensive
for large datasets. To address this challenge,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and test the hypothesis that the sample $\theta'$ based on an
approximate version of the test $u < \alpha(\theta,\theta')$. If the
approximate test cannot make a decision with sufficient confidence,
then the minibatch size is increased and the test repeats. This
process continues until a decision. The bounds depend on either a
asymptotic central limit theorem \cite{cutting_mh_2014} or a
concentration bound \cite{icml2014c1_bardenet14}, the latter requiring
direct bounds on the log likelihood ratio.  The problem with both
methods is the expense of resolving small differences.  In the worst
case, all $N$ data points may be needed, and as we show next this
worst case can occur about $\Omega(N)$ times during the performance of
$N$ tests.

Following ~\cite{icml2014c1_bardenet14}, we write the test $u < \alpha(\theta,\theta')$ in
the equivalent form $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$. Our definitions differ
from those in \cite{icml2014c1_bardenet14} by a factor of $N$ to simplify our analysis later on. 
\begin{equation}
\Lambda(\theta,\theta') = \sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta,\theta') = \log\left(u\frac{q(\theta'|\theta)p_0(\theta)}{q(\theta|\theta')p_0(\theta')}\right)
\end{equation}

To reduce computational effort, an unbiased estimate of $\Lambda(\theta,\theta')$
based on a minibatch can be used:

\begin{equation}
\Lambda^*(\theta,\theta') = \frac{N}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}

Finally, it will be convenient in what follows to define the individual components that
contribute to the sums above:

\begin{equation}\label{eq:individual_terms}
\Lambda_i(\theta,\theta') = N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}

So $\Lambda(\theta,\theta')$ is the mean of $\Lambda_i(\theta,\theta')$ over the
entire dataset, and $\Lambda^*(\theta,\theta')$ is the mean over the minibatch
of $\Lambda_i(\theta,\theta')$.

Since the minibatch contains randomly selected samples $x_i$, the
values $\Lambda_i$ are independent, IID random variables.  By the central
limit theorem, we expect $\Lambda^*(\theta,\theta')$ to be
approximately Normal.  The acceptance test then becomes a statistical
test of the hypothesis that $\Lambda(\theta,\theta') >
\psi(u,\theta,\theta')$ by establishing that
$\Lambda^*(\theta,\theta')$ is substantially larger than
$\psi(u,\theta,\theta')$.  In ~\cite{cutting_mh_2014} an asymptotic
central limit argument was used to derive this gap, while in
\cite{icml2014c1_bardenet14} a concentration bound was used. In both
cases, the resulting tests were shown to give useful reductions in
number of samples required over using the full dataset. But no
worst-case bounds were given on average number of samples needed.

We next show that for some simple distributions, the lower
bound of average number of data instances consumed for one iteration of
~\cite{cutting_mh_2014} and ~\cite{icml2014c1_bardenet14} is
$\Omega(N)$, where $N$ is the number of data points.

\subsection{An Example}
Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{N}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta)^2\right) =
  N(\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta)^2N/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta)$
and $\psi(u,\theta',\theta)$.

We would like to sample from the posterior distribution of $\theta$
which is a normal distribution centered on the sample mean, and with
variance $1/N$.  The target distribution is therefore
$\mathcal{N}(\mu,1/N)$ where $\mu$ is the sample mean, and $\mathcal{N}(\mu,\sigma^2)$ is a
normal distribution with mean $\mu$ and variance $\sigma^2$.  An efficient
proposal distribution has the same variance as the target distribution in one dimension
\cite{OptimalScaling01}, so take the proposal
$\mathcal{N}(\theta'-\theta,1/N)$, which is implemented as $q(\theta'\mid\theta) =
\phi((\theta'-\theta)N)$, where $\phi(x)$ is the Normal density
function with zero mean and unit variance. This proposal is symmetric
$q(\theta'\mid\theta)=q(\theta\mid\theta')$, and we assumed uniform
prior for $\theta$, so $\psi(u,\theta',\theta)$ reduces to $\log
u$.

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed (independent of $N$) constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta, \theta', u)$, the
  tests from \cite{icml2014c1_bardenet14} and \cite{cutting_mh_2014} consume all $N$ samples. 
\end{lemma}
Proof is given in the appendix.

Similar results can be proved for other distributions and proposals, by identifying
regions in product space $(\theta,\theta'-\theta,u)$ such that the hypothesis test
needs to separate nearly-equal values.

It follows that the accelerated M-H tests in \cite{icml2014c1_bardenet14} and
\cite{cutting_mh_2014} require at least a constant fraction $\geq c$
in the amount of data consumed per test, compared to full-dataset tests.
i.e. their speed-up is at most $1/c$.

\subsection{MCMC Posterior Inference}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

%% Daniel: for a further discussion on the 'Firefly' paper, see this blog post:
%% https://xianblog.wordpress.com/2014/04/02/firefly-monte-carlo/
%% Note: this guy also blogged about our earlier ICLR 2015 paper. He's the third author of the SAME
%% paper, which explains why he knows this stuff.

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm log}$ from now on). In \cite{Bardenet15} it was proposed to
use $\Delta^*$ in a Barker test anyway and tolerate the fixed error caused by this approximation.

@@

%% \section{Analysis}\label{sec:theory}

>>>>

\newcommand{\simiid}{\overset{\textrm{i.i.d.}}{\sim}}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}

@@

\author{
  John Canny, Haoyu Chen, Daniel Seita, Xinlei Pan \\
  University of California, Berkeley \\
  \texttt{\{canny,haoyuchen,seita,xinleipan\}@berkeley.edu}
 }

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to perform
Metropolis-Hastings tests on the full data. Previous work on reducing
the cost of Metropolis-Hastings tests yield variable data consumed per
sample, with only constant factor reductions versus using the full
dataset for each sample.  Here we present a method that can be tuned
to provide arbitrarily small batch sizes, by adjusting either proposal
step size or temperature. Our approach uses the natural noise present
in minibatch likelihood estimates to furnish the randomness in a
Metropolis-Hastings test. Our test uses the noise-tolerant Barker
acceptance test with a novel additive correction variable.  The
resulting test can be combined with minibatch proposals to yield
updates with the same complexity as a simple SGD update. In this paper
we derive the test, analyze its performance, discuss its
implementation, and present several experiments.
\end{abstract}

@@

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. We are interested primarily
in large dataset applications, where the goal is to sample a posterior
distribution $p(\theta \mid x_1, \ldots, x_N)$ of parameter $\theta$,
and where the number of data instances $N$ is large.  The
Metropolis-Hastings method (M-H) generates sample candidates from a
proposal distribution $q$ which is in general different from the
target distribution $p$, and decides whether to accept or reject them
based on an acceptance test. The acceptance test is usually a
Metropolis test~\cite{Metropolis1953,hastings70}. Conventional
Metropolis-Hastings requires all $N$ data instances to generate one
posterior sample.

Many state-of-the-art machine learning methods, and deep learning in particular,
are based on minibatch updates (such as SGD) to a model.  Minibatch updates
produce many improvements to the model for each pass over the dataset, and have
high sample efficiency. They also map very well onto hardware such as GPUs. In
contrast, M-H requires calculations over the full dataset to produce a new
sample.  Recent results from~\cite{cutting_mh_2014,icml2014c1_bardenet14}
perform approximate (bounded error) acceptance tests using subsets (minibatches)
of the full dataset. The tests depend on minibatch statistics, and on the value
of an additional uniform random variable $u$. The amount of data consumed for
each test varies significantly from one minibatch to the next, and depends on
the current sample, the proposed sample, and on the random variable $u$. By
contrast,~\cite{conf/uai/MaclaurinA14} performs exact tests but requires a lower
bound on parameter distribution across its domain.  The amount of data reduction
depends on the accuracy of this bound, and such bounds are only available for
relatively simple distributions.

Here we derive a new test which incorporates the variability in
minibatch statistics as {\em a natural part of the test}. Because of
this, the amount of data required for each test is fixed with high
probability and in expected value. We use a Barker test function
\cite{Barker65} rather than a Metropolis test, which makes our test
naturally error tolerant. The idea of using a noise-tolerant test
using Barker's test function was suggested but not explored
empirically in \cite{Bardenet15} section 6.3. But the asymptotic test
statistic CDF and the Barker function are different, which leads to
fixed errors for the approach in \cite{Bardenet15}. Here we show that
the difference between the distributions can be corrected with an
additive random variable. This leads to a test which is fast, and
whose error can be made arbitrarily small.

@@

\item Increase the minibatch size. Log acceptance variance scales as $1/k$ vs the
  minibatch size $k$. Increased minibatch size also reduces the error rate for
  the test. {\color{blue} Daniel: I'm not sure if I'd mention this here, because
  before this list, we talk about reducing the data, and then suddenly we're
  talking about increasing it.}

@@

Its worth discussing at this point the typical goals of
M-H sampling on very large datasets.  By the Bernstein-von Mises
Theorem, the posterior distribution of the parameter $\theta$ for a
Bayesian inference task is asymptotically normal, and has variance
that scales inversely with the number of data samples $N$. This mode
is extremely sharp for large datasets, which may contain millions or
billions of samples. Simply sampling from this distribution is one
application, but an efficient proposal distribution
\cite{OptimalScaling01} has similar variance to the target
distribution and will diffuse to it extremely slowly from an
initialization value which is (likely to be) many standard deviations
away. If there are any other strong modes, it is very likely for the
sampler to find one of them and become trapped in it when run at the
normal distribution temperature ($T=1$). A common solution is to anneal
the sampler, running first at high temperature (scaling log
likelihoods by $1/T$) which flattens the likelihood landscape.  This is
turn reduces the variance of the log acceptance probability and allows
our acceptance test to be applied.

A second question concerns step size. Once we have fixed temperature,
our variance constraint implies that we have to trade-off proposal
step size $s$ and batch size $b$ ($b \propto p^2$). i.e. we can make
many small steps, or one large step, with a given batch of data. One
of the primary drivers of this work is our belief in the value of
small steps. For applications to neural networks or other models where
the posterior is multimodel, posterior inference is arguably a search
process. Covering the search space densely with small steps is much
more valuable than few sparse steps toward the nearest optimum. In
this mode, Metropolis-Hastings can be used in similar fashion to
Stochastic Gradient Descent. The goal in SGD is to make gradual
progress to a posterior mode with each step, taking small steps so
that the cumulative displacement has progressively lower variance. A
substantial part of the computational work of MCMC on massive datasets
will be similarly in reaching a stationary distribution, which really
means finding a deep posterior mode. Taking noisy small steps will
nevertheless make steady progress to a posterior mode since their bias
is in that direction. We demonstrate this behavior in our logistic regression
experiments.

@@

\begin{itemize}
\item We develop a new, more efficient (in samples per test) minibatch acceptance test with quantifiable error bounds. The test
  uses a novel additive correction variable to implement a Barker test based on minibatch mean and variance. 
\item We analyze the test for accuracy and speed.
\item We compared performance of our new test and prior approaches on several datasets. We demonstrate the test is several orders
  of magnitude more efficient than prior work measured as data consumed per test, and that it does not suffer
  from long-tailed minibatch sizes (up to the dataset size). 
\end{itemize}

@@

In the Metropolis-Hastings method~\cite{gilks1996markov,brooks2011handbook}, a
difficult-to-compute probability distribution $p(\theta)$ is sampled using a
Markov chain $\theta_1,\ldots,\theta_n$. The sample $\theta_{t+1}$ at time $t+1$
is generated using a candidate $\theta'$ from a (simpler) proposal distribution
$q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test
is usually a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
    \alpha(\theta_t,\theta') = \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)} \wedge 1
\end{equation}
where $a \wedge b$ denotes $\min(a,b)$.  With probability
$\alpha(\theta_t,\theta')$, we accept $\theta'$ and set $\theta_{t+1} =
\theta'$, otherwise set $\theta_{t+1}=\theta_t$.  The test is often implemented
with an auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison
$u<\alpha(\theta_t,\theta')$; here, $\mathcal{U}(a,b)$ denotes the uniform
distribution on the interval $[a,b]$.  For notational simplicity, we drop the
subscript $t$ for the current sample $\theta_t$ and denote it as $\theta$.

The acceptance test guarantees detailed balance, which means
$p(\theta)p(\theta'\mid\theta) = p(\theta')p(\theta \mid\theta')$,
where $p(\theta'\mid\theta)$ is the probability of a transition from state
$\theta$ to $\theta'$. Here $p(\theta'\mid\theta) =
q(\theta'\mid\theta)\alpha(\theta,\theta')$ so the detailed balance equation
becomes
\begin{equation}\label{eq:detailed_balance2}
    p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
\end{equation}
This condition, together with ergodicity, guarantees that the Markov chain has a
unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution
for $\theta$ based on some observed data, so the target distribution is
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}\label{eq:acceptance_probability}
    \alpha(\theta,\theta') = 
    \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
    \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
    \wedge 1
\end{equation}
where $p_0(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the probabilities
of the observations. Computing samples this way requires the use of all $N$
training data points, but this is very expensive for large datasets. To address
this challenge,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform approximate
Metropolis-Hasting tests using sequential hypothesis testing. During each
iteration, they start with a small minibatch of data and test the hypothesis
that the sample $\theta'$ should be accepted based on an approximate version of
the test $u < \alpha(\theta,\theta')$. If the approximate test cannot make a
decision with sufficient confidence, then the minibatch size is increased and
the test repeats. This process continues until a decision. The bounds depend on
either a asymptotic central limit theorem \cite{cutting_mh_2014} or a
concentration bound \cite{icml2014c1_bardenet14}. The latter requires direct
bounds on the log likelihood ratio, which for general distributions requires
knowing $p(x_i \mid \theta)$ and $p(x_i \mid \theta')$ for all $N$ samples. In
addition, both methods suffer the drawback of resolving small log likelihood
ratio differences between the minibatch and full batch versions.  In the worst
case, all $N$ data points may be needed, and we soon show that this worst case
can occur about $\Omega(N)$ times during the performance of $N$ tests.

Following~\cite{icml2014c1_bardenet14}, we write the test $u <
\alpha(\theta,\theta')$ in the equivalent form $\Lambda(\theta,\theta') >
\psi(u,\theta,\theta')$, where\footnote{Our definitions differ from those in
\cite{icml2014c1_bardenet14} by a factor of $N$ to simplify our analysis later.}
\begin{equation}
\Lambda(\theta,\theta') = \sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta,\theta') = \log\left(u\frac{q(\theta'|\theta)p_0(\theta)}{q(\theta|\theta')p_0(\theta')}\right)
\end{equation}
To reduce computational effort, an unbiased estimate of $\Lambda(\theta,\theta')$
based on a minibatch can be used:
\begin{equation}
\Lambda^*(\theta,\theta') = \frac{N}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}
Finally, it will be convenient for our analysis to define the individual
components that contribute to the sums above:
\begin{equation}\label{eq:individual_terms}
\Lambda_i(\theta,\theta') = N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}
Thus, $\Lambda(\theta,\theta')$ is the mean of $\Lambda_i(\theta,\theta')$ over
the entire dataset, and $\Lambda^*(\theta,\theta')$ is the mean of
$\Lambda_i(\theta,\theta')$ over its minibatch.

Since minibatches contains randomly selected samples $x_i$, the values
$\Lambda_i$ are independent, identically distributed (iid) random variables.
{\color{blue} Daniel: should we say sampling with or without replacement? Right
now I think the code does it without replacement (but \emph{with} replacement
means it's closer to the CLT, right?} By the Central Limit Theorem, we expect
$\Lambda^*(\theta,\theta')$ to be approximately Gaussian. The acceptance test
then becomes a statistical test of the hypothesis that
$\Lambda(\theta,\theta')>\psi(u,\theta,\theta')$ by establishing that
$\Lambda^*(\theta,\theta')$ is substantially larger than
$\psi(u,\theta,\theta')$.  In~\cite{cutting_mh_2014} an asymptotic central limit
argument was used to derive this gap, while in~\cite{icml2014c1_bardenet14} a
concentration bound was used. In both cases, the resulting tests were shown to
give useful reductions in number of samples required over using the full
dataset, but there were no worst-case bounds given on the average number of
samples needed per iteration.

We next show that for a simple distribution, the lower bound of average
number of data instances consumed for one iteration of~\cite{cutting_mh_2014}
and~\cite{icml2014c1_bardenet14} is $\Omega(N)$, where $N$ is the number of data
points.

\subsection{A Worst-Case Gaussian Example}\label{ssec:gaussian_example}
Consider a Gaussian model where $x_1,\ldots,x_N \simiid \mathcal{N}(\theta,1)$
with a known variance $\sigma^2=1$, true mean $\theta=0.5$, and a uniform prior
on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{N}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta)^2\right) =
  N(\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.  Since
the $x_i$ have unit variance, their mean has variance $1/b$, and the variance of
$\Lambda^*(\theta,\theta')$ is $\sigma^2(\Lambda^*) = (\theta'-\theta)^2N^2/b$.
In order to pass a hypothesis test that $\Lambda > \psi$, there needs to be a
large enough gap (several $\sigma(\Lambda^*)$) between
$\Lambda^*(\theta',\theta)$ and $\psi(u,\theta',\theta)$. {\color{blue}
Daniel: I changed the variance to be $(\theta'-\theta)^2N^2/b$; before it had
$N$ instead of $N^2$.}

Our goal is to sample from the posterior $p(\theta \mid x_1,\ldots,x_N)$, which
is a normal distribution $\mathcal{N}(\mu, 1/N)$ centered on the sample mean
$\mu$, and with variance $1/N$. In one dimension, an efficient proposal
distribution has the same variance as the target
distribution~\cite{OptimalScaling01}, so we use the proposal
$\mathcal{N}(\theta'-\theta,1/N)$, which is implemented as
$q(\theta'\mid\theta)=\phi((\theta'-\theta)N)$, where $\phi(x)$ is the Normal
density function with zero mean and unit variance. This proposal is symmetric
$q(\theta'\mid\theta)=q(\theta\mid\theta')$, and since we assumed a uniform
prior on $\theta$, we see that $\psi(u,\theta',\theta)$ reduces to $\log u$. Our
worst-case scenario is specified in Lemma~\ref{lem:worst_case}.
{\color{blue} Daniel: this might be a silly question, but why do we assume the
proposal has mean $\theta'-\theta$? I thought it would have a mean of $\theta$
because that is the current sample and the Gaussian is centered at that point?}

\begin{lemma}\label{lem:worst_case}
    For the model in Section~\ref{ssec:gaussian_example}, there exists a fixed
    (independent of $N$) constant $c$ such that with probability $\geq c$ over
    the joint distribution of $(\theta, \theta', u)$, the tests
    from~\cite{cutting_mh_2014,icml2014c1_bardenet14} consume all $N$ samples. 
\end{lemma}

\begin{proof}
See Appendix~\ref{app:worst_case_proof}.
\end{proof}

Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta'-\theta,u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated M-H tests in~\cite{cutting_mh_2014,icml2014c1_bardenet14} require at
least a constant fraction $\geq c$ in the amount of data consumed per test
compared to full-dataset tests, so their speed-up is at most $1/c$.

\subsection{MCMC Posterior Inference}
There is a separate line of MCMC work drawing principles from statistical
physics. By viewing random variables as particles in a system, one can apply
Hamiltonian Monte Carlo (HMC)~\cite{mcmc_hamiltonian_2010} methods which
generate high acceptance \emph{and} distant proposals when run on full batches
of data. Recently Langevin Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has
been applied to Bayesian estimation on minibatches of data. This simplified
dynamics uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on proposal
step size is severe, and the state space exploration reduces to a random walk.
Full minibatch HMC for minibatches was recently described in~\cite{sghmc_2014}
which allows momentum-augmented proposals with larger step sizes. However, step
sizes are still limited by the need to run accurately without MH tests.  Our
work opens the door to applying those methods with much more aggressive step
sizes without loss of accuracy. We demonstrate this in Section ... {\color{blue}
Daniel: do we ever demonstrate this? I guess another issue is that this implies
we can use larger step sizes, but earlier we \emph{also} argue that we need
\emph{small} step sizes (to reduce variance).}

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$,
we can write
\begin{equation}\label{eq:relationship}
    \Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is
approximately normal with variance $\sigma^2(\Delta^*)$.  But to perform the
test in Equation~\ref{eq:the_exact_test} we want $\Delta + X$ for a logistic
random $X$ (call it $X_{\rm log}$ from now on). In \cite{Bardenet15} it was
proposed to use $\Delta^*$ in a Barker test anyway and tolerate the fixed error
caused by this approximation.

@@
